---
title: Analysis for 'Does persistent colour preference constrain behavioural flexibility in Heliconius butterflies?'
output:
  bookdown::html_document2:
    includes:
      in_header: docs/header.html
      after_body: docs/footer.html      
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    split_by: section
    css: styles.css

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = 'docs/analysis')
  })
---

<p class="author-name">M. Wyatt Toure</p>

<p>e-mail: `wyatt.toure@gmail.com`<br></p>

```{r library-prep, include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(cowplot)
library(lme4)
library(lmerTest)
library(effects)
library(emmeans)
library(ggpubr)
library(rptR)
source("R/tidy-exploration-data.R")
source("R/tidy-learning-data.R")
```

*** 

## Brief Overview

This page details the analysis performed for 'untitled Heliconius exploration
project'. I walk through all the data preparation steps and models here and
explain my reasoning for analysis decisions. If you are only interested in the
model outputs you can go straight to the [Models](#models) section. The code to
reproduce this analysis are embedded within the code to generate this page which
is in the
[`analysis.Rmd`](https://github.com/wyatt-toure/heliconius-exploration/blob/main/analysis.Rmd)
file. The data underlying this analysis are available in the
[`/data`](https://github.com/wyatt-toure/heliconius-exploration/blob/main/data)
directory. Instructions on how to run the analyses as well as metadata for the
variables can be found in
[README.md](https://github.com/wyatt-toure/heliconius-exploration#how-to-reproduce-the-results).

***

## Background

One of the things I noticed while I was doing an experiment on Heliconiini
time-based learning was that some individuals did not switch colours very much
between time sessions, even though the flower was actively punished. This can be
seen below where I plot the proportion of correct choices made during all of
training.

```{r flexibility-plot, echo=FALSE, warning=FALSE, fig.width=10, fig.cap="The proportion of correct choice an individual made during the morning and the afternoon throughout all of training. Lines connect individual acoss time periods. "}
learning_data <- read.csv("data/heliconiini-learning-data.csv")

ggplot(learning_data, aes(
  x = time.of.day,
  y = (n.correct.training / (n.correct.training + n.incorrect.training))
)) +
  theme_cowplot() +
  geom_point(color = "#444444") +
  geom_line(aes(group = id), color = "#bab0ac", alpha = 0.8) +
  ylab("Proportion correct training") +
  xlab("Time of day") +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold")
  )
```

Some of these lines have steeper slopes than others. Since the correct choice in
the morning is incorrect in the afternoon, this means that steep slopes indicate
individuals who stuck with their 'favourite' colour (stubborn). The less steep
the slope, the more flexible an individual was. I was interested in whether
these steep sloped individuals were generally less exploratory/flexible than
their comparatively less stubborn counterparts.

For this experiment we wanted to understand the following:

1.  Are stubborn (steep sloped) individuals less likely to sample a novel
    colour ([Model 1](#model-1))?
2.  Does the stubborn group remain stubborn during exploration trials? Do they
    always stick with their favourite colour ([Model 3](#model-3))?

To get at this I gave butterflies exploration trials where familiar colours
(yellow or purple) were paired with novel colours which the butterflies had
never seen before.

***

## Data preparation 

### Exploration assay data

The raw data for the exploration experiment are in the form of individual
choices associated with an individual ID as well as the cage identity and the
two colours present in the assay. 

```{r data-prep}
## Read in data
exploration_data <- read.csv("data/heliconius-exploration-data-raw.csv")

## ID to factor
exploration_data$id <- as.factor(exploration_data$id)
```

#### Standardise combination names 

Here I standardise the levels of the variable `combo` so that the colour
combinations are all in the same order with the familiar colour first and the
novel colour second.

```{r data-prep2}
## Renaming factor levels
## Old name is on the left, new name is on the right
exploration_data$combo <- recode_factor(exploration_data$combo,
                                                    "blue-purple" = 'purple-blue',
                                                    "blue-yellow" = 'yellow-blue',
                                                    "dblue-purple" = 'purple-blue',
                                                   "lblue-yellow" = 'yellow-lblue',
                                                    "orange-purple" = 'purple-orange',
                                                    "pink-purple" = 'purple-pink',
                                                    "pink-yellow" = 'yellow-pink',
                                                    "red-purple" = 'purple-red',
                                                   "red-yellow" = 'yellow-red'
                                                   )
```

#### Summarise raw counts

I tabulated the raw data into their totals using a custom function
`summarise_exploration_data()`. You can see the code for this function [here]().

```{r data-prep3}
## Applying summarise exploration data function
exploration_data_summarised <- summarise_exploration_data(exploration_data)
```

#### Assign time of day and rewarding colours

Next I needed to assign the time of day for the trials as well as their original
session rewarding colours from the learning experment. This was done with a
series of cage-specific `case_when()` and `mutate()` statements. This section is
quite long so I have folded the code, you can click the show code tab to see the
code.

```{r data-prep4}
# Assigning time of day to assays
exploration_data_summarised <- exploration_data_summarised %>% 
  mutate(
    time.of.day = case_when(
      combo == "yellow-blue" & cage == "1A" ~ "1morning",
      combo == "yellow-lblue" & cage == "1A" ~ "1morning",
      combo == "purple-red" & cage == "1A" ~ "2afternoon",
      combo == "purple-pink" & cage == "1A" ~ "2afternoon",
      
      combo == "purple-blue" & cage == "2A" ~ "1morning",
      combo == "yellow-lblue" & cage == "2A" ~ "1morning",
      combo == "yellow-red" & cage == "2A" ~ "2afternoon",
      combo == "purple-orange" & cage == "2A" ~ "2afternoon",
      
      combo == "yellow-blue" & cage == "2B" ~ "1morning",
      combo == "purple-orange" & cage == "2B" ~ "1morning",
      combo == "purple-red" & cage == "2B" ~ "2afternoon",
      combo == "yellow-pink" & cage == "2B" ~ "2afternoon",
      
      combo == "yellow-lblue" & cage == "3A" ~ "1morning",
      combo == "purple-blue" & cage == "3A" ~ "2afternoon",
      
      combo == "yellow-orange" & cage == "1B" ~ "1morning",
      combo == "purple-red" & cage == "1B" ~ "1morning",
      combo == "purple-pink" & cage == "1B" ~ "2afternoon",
      combo == "yellow-blue" & cage == "1B" ~ "2afternoon",
      
      combo == "yellow-blue" & cage == "3B" ~ "1morning",
      combo == "purple-pink" & cage == "3B" ~ "1morning",
      combo == "purple-red" & cage == "3B" ~ "2afternoon",
      combo == "yellow-orange" & cage == "3B" ~ "2afternoon"
    )
  )

# Assigning original rewarding colours for the time period based on the cage
exploration_data_summarised <- exploration_data_summarised %>% 
  mutate(
    session.reward.colour =
      case_when(
        cage == "1A" & time.of.day == "1morning" ~ "yellow",
        cage == "1A" & time.of.day == "2afternoon" ~ "purple",
        
        cage == "2A" & time.of.day == "1morning" ~ "purple",
        cage == "2A" & time.of.day == "2afternoon" ~ "yellow",
        
        cage == "2B" & time.of.day == "1morning" ~ "yellow",
        cage == "2B" & time.of.day == "2afternoon" ~ "purple",
        
        cage == "3A" & time.of.day == "1morning" ~ "yellow",
        cage == "3A" & time.of.day == "2afternoon" ~ "purple",
        
        cage == "1B" & time.of.day == "1morning" ~ "purple",
        cage == "1B" & time.of.day == "2afternoon" ~ "yellow",
        
        cage == "3B" & time.of.day == "1morning" ~ "yellow",
        cage == "3B" & time.of.day == "2afternoon" ~ "purple"
      )
  )
```


```{r setting-colour-values, echo=FALSE}
# This code uses tableau colour palettes to create a vector a hex code
# colour values
red <- "#C0392B"
purple <- "#8E44AD"
yellow <- "#F1C40F"
orange <- "#DC7633"
blue <- "#2980B9"
pink <- "#F08080"
```

### Learning experiment data

The learning data come from 'Heliconiini butterflies can learn
time-dependent reward associations' which also has a GitHub repository called
[`butterfly-learning`](https://wyatt-toure.github.io/butterfly-learning/). In
that experiment I also looked at the species *Dryas iulia* but they were not
assayed for this experiment so I filter the data down to only *Heliconius*
species.

```{r learning-data-prep}
## Read in data
learning_data <- read.csv("data/heliconiini-learning-data.csv")

## Retrieve only Heliconius species by excluding Dryas
learning_data <- learning_data %>% filter(species != "dryas")
```

#### Get actual colours of choices

The choice data in this sheet are coded with regards to whether they are correct
or incorrect as well as whether they are the 'colour rewarded in the morning' or
'colour rewarded in the afternoon'. I needed to get the actual colours of the
choices so I converted the choices based on what the correct reward colour was
for the session using the `session.reward.colour` variable.

```{r getting-true-colour-choices}
# Converting training choices to their colour versions
learning_data <- learning_data %>%
  mutate(
    n.purple.training =
      case_when(
        session.reward.colour == "purple" ~ n.correct.training,
        TRUE ~ n.incorrect.training
      )
  ) %>%
  mutate(
    n.yellow.training =
      case_when(
        session.reward.colour == "yellow" ~ n.correct.training,
        TRUE ~ n.incorrect.training
      )
  )
```

#### Stubborness index calculation 

I created a stubbornness index which was created by the following formula

 | correctness in afternoon - correctness in morning |

This means that a butterfly that was stubborn will have stubbornness index
closer to 1 and a flexible butterfly will have a stubbornness index closer to 0.

```{r}
# Creating proportional preference measure for learning test and training data
learning_data <- learning_data %>%
  mutate(
    morning.colour.pref.prop =
      n.morning.colour.test / (n.morning.colour.test + n.afternoon.colour.test)
  ) %>%
  mutate(
    training.correctness.prop =
      n.correct.training / (n.correct.training + n.incorrect.training)
  )

# Getting the shift in colour preference during testing in learning experiment
learned_shift_in_colour_pref_data <- learning_data %>%
  group_by(id) %>%
  summarise(
    shift.in.morning.colour.preference =
      abs(last(morning.colour.pref.prop) - first(morning.colour.pref.prop))
  )

# Getting the shift in colour preference during training in learning experiment
training_shift_in_colour_pref_data <- learning_data %>%
  group_by(id) %>%
  summarise(
    training.stubborness.index =
      abs(last(training.correctness.prop) - first(training.correctness.prop))
  )
```

Finally we join the data sets

```{r}
full_data <-
  left_join(exploration_data_summarised, learned_shift_in_colour_pref_data, 
            by = "id") %>%
  left_join(training_shift_in_colour_pref_data, 
            by = "id")
```

***

## Models {#models}

Note for the time being I just have the raw model summary outputs. I will format
the outputs when I have more time later on.

### Model 1 - Are stubborn individuals less likely to sample a novel colour? {#model-1}

The first model I tried is a simple model with just the proportional preference
for a novel colour and the effect of the training stubbornness index. This model
asks whether the proportion of novel colour choices is associated with how
stubborn an individual was during training for the learning experiment.

- Response variable: `cbind(n.novel.colour, n.familiar.colour)`
- Fixed effect: `training.stubborness.index`
- Random effect: `id`
- Random effect: `cage`
- Random effect: `combo`

`id` is nested within `cage`.

```{r}
exploration_model <- glmer(
  cbind(n.novel.colour, n.familiar.colour) ~ 
    training.stubborness.index + (1 | cage / id) + (1 | combo),
  data = full_data,
  family = "binomial"
)
```

```{r}
summary(exploration_model)
```


What I find is that there is no significant effect though, the effect size is in
the expected direction. There is a non-significant relationship between the
proportion of novel colour choices and the training stubbornness index whereby
more stubborn individuals were less likely to sample a novel colour. We also see
that there is more variation among individuals than among cages so individuals
are not likely to be copying each other. 

The largest amount of variation by far is in the `combo` random effect, which
makes sense since there is variation in what colours butterflies were trained to
and for what time periods. I will have to address this variation in subsequent
models.

### Model 2 - Were flexible individuals at test more likely to sample new colours? {#model-2}

In this next model I ask whether the proportion of novel colour choices is influenced by how flexible an individual was during the final test for the learning experiment.

```{r}
exploration_model2 <- glmer(
  cbind(n.novel.colour, n.familiar.colour) ~ 
    shift.in.morning.colour.preference + (1 | cage / id) + (1 | combo),
  data = full_data,
  family = "binomial"
)
```

```{r}
summary(exploration_model2)
```

There is no significant effect. Looking at the direction of the effect reveals
that individuals which did not shift their colour preference at test sampled
less novel colours. This might be a consequence of learning the task more
effectively. Perhaps they stuck with familiar colours because they were more
responsive to training. However, we probably should not read too much into a
non-significant result.

### Model 3 - Does a stubborn individual stick with their favourite colour during exploration assays? {#model-3}

Our stubborn individuals may have different colours as their favourite colour.
To investigate whether this is maintained over exploratory behaviour assays I
determined what the identity of the colour an individual made the vast majority
of choices for was. To determine what a butterfly's favourite colour was, I took
the sum of all choices they had during the 10 day training period and assigned
them as having either purple or yellow as their favourite colour.

*Note*, this is probably a crude way of doing this for now during my exploratory
investigations. I have plans to also make a metric that incorporates the
magnitude of the preference *i.e.*, 51 purple 50 yellow and 100 purple 1 yellow
shouldn't be seen as the same.

```{r}
# Get total number of purple and yellow choices
total_training_choices <- learning_data %>%
  group_by(id) %>%
  summarise(across(c(n.purple.training, n.yellow.training), sum))

# Determine which colour has more choices
total_training_choices <- total_training_choices %>% 
  mutate(
    favourite.colour = case_when(
      n.yellow.training > n.purple.training ~ "yellow",
      n.purple.training > n.yellow.training ~ "purple"
    )
  )

# Extract favourite colours and ids
favourite_colour_matrix <- total_training_choices %>% 
  select(id, favourite.colour)

# Join data frames
full_data <- left_join(full_data, favourite_colour_matrix,
    by = "id"
  )

# Adding the identity of the familiar and novel colour as a variable
full_data <- full_data %>% mutate(
      familiar.colour =
        case_when(
          combo == "purple-blue" ~ 'purple',
          combo == "yellow-blue" ~ 'yellow',
          combo == "yellow-lblue" ~ 'yellow',
          combo == "purple-orange" ~ 'purple',
          combo == "purple-pink" ~ 'purple',
          combo == "yellow-pink" ~ 'yellow',
          combo == "purple-red" ~ 'purple',
          combo == "yellow-red" ~ 'yellow',
          combo == "yellow-orange" ~ 'yellow'
        )) %>%
    mutate(
      novel.colour =
        case_when(
          combo == "purple-blue" ~ 'blue',
          combo == "yellow-blue" ~ 'blue',
          combo == "yellow-lblue" ~ 'lblue',
          combo == "purple-pink" ~ 'pink',
          combo == "yellow-pink" ~ 'pink',
          combo == "purple-orange" ~ 'orange',
          combo == "purple-red" ~ 'red',
          combo == "yellow-red" ~ 'red',
          combo == "yellow-orange" ~ 'orange'
        ))
```

I asked whether the identity of the favourite colour and the identity of the
familiar colour influences the proportion of novel colour choices an individual
makes. Importantly this might vary with what colour was originally rewarding
during the time of day the assay takes place so I have included
`session.reward.colour` as a factor in the model.

```{r}
exploration_model3 <- glmer(
  cbind(n.novel.colour, n.familiar.colour) ~ 
    favourite.colour * familiar.colour * session.reward.colour + (1 | cage / id) + (1 | combo),
  data = full_data,
  control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)),
  family = "binomial"
)
```

```{r}
broom::tidy(exploration_model3)
```

```{r}
emmip(exploration_model3,
  favourite.colour ~ familiar.colour:session.reward.colour,
  CIs = TRUE,
  plotit = FALSE,
  type = "response"
) %>% select(-df, -tvar, -xvar) %>% kable()
```

**Note** yvar is the proportional preference for novel colours so lower values
mean more choices for the familiar colour and higher values mean more choices or
the novel colour.

For now I seem to be seeing the exact opposite of what I expected. You would
expect that if purple is your favourite colour, purple is the familiar colour,
and purple is the original session reward colour that a butterfly would make
more familiar than novel object choices but we don't see this same with if you
replace purple with all yellows.

Glancing at the numbers seems to suggest that you see the most familiar choices
when the session reward colour is opposite of their favourite colour and the
familiar colour is the favourite colour. That is we see highest preferences or
familiar colours when

- favourite colour = yellow | familiar colour = yellow | session reward colour = purple
- favourite colour = purple | familiar colour = purple | session reward colour = yellow

One explanation could be that for those that have their favourite colour
presented as the familiar colour in a session, it does not matter what the
session rewarding colour, they go for the familiar colour regardless. Put
another way, if I see my favourite colour is here (it is the familiar colour)
then I don't care what the conditions are under—i'm going to make familiar
colour choices.

Still would need to explain why we see more novel choices when the familiar
colour, favourite colour and session reward colour are all in agreement. Perhaps
variation due to colour combinations?

Something that is interesting is that when the familiar colour is yellow and the
favourite colour is yellow and the session rewarding colour is yellow, the
preference for novelty is lower than if the familiar colour is purple as your
would expect. </br>
However, when the familiar colour is purple and the favourite
colour is purple and the session reward colour is purple, the preference for
novelty is higher than if the familiar colour is yellow, which is opposite to
what you would expect. 

Will have to dig deeper into what colours are paired with what. Maybe purple is
systematically paired with more attractive colours. Looking down at the plot in
the plot graveyard, this might be the case.

```{r fig.width=10, echo=FALSE, fig.cap="Proportion of novel colour choices for butterflies separated by the favourite colour of the individual and the identity of the familiar colour. Data are means ± SE"}
# # Getting estimated marginal means and confidence intervals
# exploration_model3_emmeans_data <- emmip(exploration_model3,
#   favourite.colour ~ familiar.colour,
#   CIs = TRUE,
#   plotit = FALSE,
#   type = "response"
# )
# 
# # Drop_nas from favourite colour 
# plot_data <- full_data %>% 
#   filter(favourite.colour == "purple" | favourite.colour == "yellow")
# 
# ggplot(exploration_model3_emmeans_data,
#          aes(
#            x = favourite.colour,
#            y = yvar,
#            color = familiar.colour
#          )
#   ) +
#   scale_color_manual(values = c(purple,yellow)) +
#     theme_cowplot() +
#     geom_jitter(data = plot_data,
#                 aes(x = favourite.colour, 
#                     y = (plot_data$n.novel.colour/(plot_data$n.novel.colour + plot_data$n.familiar.colour)),
#                     color = familiar.colour),
#                 width = 0.05,
#                 alpha = 0.2
#     ) +
#     geom_point(
#       shape = 15,
#       size = 3.5,
#       position = position_dodge(width = 0.1)
#     ) +
#   geom_errorbar(
#       aes(ymax = yvar+SE, 
#           ymin = yvar-SE), 
#       width = 0.05,
#       position = position_dodge(width = 0.1)
#     ) +
#   theme(
#     legend.position = "right",
#     axis.text = element_text(size = 12),
#     axis.title = element_text(size = 14, face = "bold")
#   ) +
#   ylab("Novel colour choice proportion")
```

```{r fig.width=10, echo=FALSE, fig.cap="This plot separates the data into 2 panels by session rewarding colour. Data are means ± SE"}
# Getting estimated marginal means and confidence intervals
exploration_model3_emmeans_data <- emmip(exploration_model3,
  favourite.colour ~ familiar.colour:session.reward.colour,
  CIs = TRUE,
  plotit = FALSE,
  type = "response"
)

# Drop_nas from favourite colour 
plot_data <- full_data %>% 
  filter(favourite.colour == "purple" | favourite.colour == "yellow")

ggplot(exploration_model3_emmeans_data,
         aes(
           x = favourite.colour,
           y = yvar,
           color = familiar.colour
         )
  ) +
  scale_color_manual(values = c(purple,yellow)) +
    theme_cowplot() +
    geom_jitter(data = plot_data,
                aes(x = favourite.colour, 
                    y = (plot_data$n.novel.colour/(plot_data$n.novel.colour + plot_data$n.familiar.colour)),
                    color = familiar.colour),
                width = 0.05,
                alpha = 0.2
    ) +
    geom_point(
      shape = 15,
      size = 3.5,
      position = position_dodge(width = 0.1)
    ) +
  geom_errorbar(
      aes(ymax = yvar+SE, 
          ymin = yvar-SE), 
      width = 0.05,
      position = position_dodge(width = 0.1)
    ) +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold")
  ) +
  facet_wrap(~session.reward.colour) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  ylab("Novel colour choice proportion")
```



***

## Plot Graveyard

Here I have plots of things I want to visualize, nothing final present here.



```{r general-trend-plot, echo=FALSE, fig.width=10, fig.cap="Data are means ± SE."}
# Colour values for plot 
color.values <- c(blue, blue, orange, pink, purple, red, yellow)

# Plot code
ggplot(
  expl_data_wide,
  aes(
    x = choice,
    y = n,
    color = choice
  )
) +
  geom_jitter(alpha = 0.4, width = 0.05) +
  scale_color_manual(values = color.values) +
  theme_cowplot() +
  theme(legend.position = "none") +
  stat_summary(
    geom = "point",
    fun = "mean",
    size = 3.5,
    shape = 15
  ) +
  stat_summary(
    geom = "errorbar",
    fun.data = "mean_se",
    position = position_dodge(width = 0),
    width = 0.3
  ) +
  facet_grid(~combo,
    scales = "free",
    space = "free"
  ) + 
  ylab("Number of choices") +
  xlab("Colour choice")
```

